### 2016-02-21 p00-ch00-intro/intro.tex
Apply discussion comments on intro

A considerable amount of research has been carried out to mix object-oriented programming languages with the actor model~\cite{philippsen2000survey}. 
Language extensions, libraries, and even new programming languages are the natural outcomes of such research.
The more mainstream the language, the more research and development are available in the intersection of object orientation and actor model.
For example, \cite{actor_frameworks_jvm:agha} presents a comparative analysis of such research for Java and JVM languages.

The rapid increase of computational power added a new aspect to create a triangle: object-oriented, actor model, and \emph{concurrency}.
Concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
Scheduling of object messages in threads of executions is one of main challenges.
In addition, in the presence of scheduling and thread interleaving, it is important that objects are able to protect their state from concurrent modifications for correctness properties.

The core contributions of this thesis target at the intersection of object 
orientation, actor model, and concurrency.
We choose Java as the main programming language and as one of the mainstream 
object-oriented languages. 
We formalize a subset of Java and its concurrency API~\cite{jsr166} to 
facilitate formal verification and reasoning about it.
We create an abstract mapping from a concurrent-object modelling language, 
ABS~\cite{johnsen2012abs}, to programming semantics of Java concurrency. 
We provide the formal semantics of the mapping and runtime properties of 
the concurrency layer including deadlines and scheduling policies.
We provide an implementation of ABS concurrency layer as a Java API library 
and framework utilizing the latest langauge additions 
in Java 8~\cite{jsr335:lambda:translation}.
We design and implement a runtime monitoring framework, JMSeq, to verify the
sequenced execution of methods through code annotatinon in JVM. 
In addition, we design a large-scale monitoring system as a real-world 
application; the monitoring system is built with actors and ABS concurrency 
with formal semantics of ABS composed with schedulability 
analysis~\cite{fersman2007task}.
We elaborate on the elements of this research in the following.

ABS~\cite{johnsen2012abs,hahnlehjlssw11} is a modelling language for concurrent objects and distributed systems.
ABS is based on the semantics of the actor model~\cite{actors:agha}.
ABS language provides a set of core features including algebraic data types (ADT), a functional layer, and a concurrency layer.
The ABS concurrency layer uses co-operative scheduling of active objects~\cite{creol:broch:owe}.
ABS semantics is completely expressed in structural operational semantics~\cite{plotkin:sos}.
This allows ABS models to take advantage of various verification methods and static analysis techniques.
% 
Co-operative scheduling in ABS has been additionally extended for real-time scheduling taking priorities and time constraintsinto account~\cite{bjork2013:rtabs,johnsen2012modeling}.
All the above characteristics make the ABS language an attractive choice if it
can be used as a programming language at industrial and business scale.
However, ABS has so far remained mainly as a modelling language.
\TODO{Check with Frank.}

One core challenge is to combine ABS semantics (e.g. concurrency layer) in a
programming language.
Considering the mainstream programming languages
\footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}, Java~\cite{gosling2000java} is one of the most commonly-used.
Since Java 1.5 and the rise of the concurrency API in Java~\cite{jsr166}, substantial effort is spend by the community on how concurrency and distribution can be improved in Java, along the lines of the actor model, as provided by ABS.
However, not all the development on Java libraries and frameworks are based on a rigorous formal semantics. 
Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.

% We deliver ABS concurrency layer in form of a Java library API.
% However, 
The Java language specification (JLS)~\cite{gosling2000java} clearly concludes that Java language is not comptabile and ready to be extended with a functional language that supports algebraic data types.
This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
The Scala language~\cite{odersky2004scala} is a result of the research. It is a dynamic, functional, and object-oriented language on JVM.
However, these extensions and libraries lack formal semantics in the context
of concurrency and actor model.

We primarily focus on ABS concurrency layer and deliver its semantics as a Java API.
We argue that the advantage of a correct semantic translation from ABS concurrency layer to its Java equivalent constructs overcomes the additional support of functional layer with \emph{partial} formal semantics.
Our approach enables programmers to model their Java systems in terms of ABS concurrency semantics and then create a correct mapping and implementation from ABS to Java using an API.
In addition, we aim to facilitate ways to actually program ABS models taking advantage of the standard Java libraries.

>>>>

A considerable amount of research has been carried out to mix object-oriented programming languages with the actor model~\cite{philippsen2000survey}. 
Language extensions, libraries, and even new programming languages are the natural outcomes of such research.
For example, \cite{actor_frameworks_jvm:agha} presents a comparative analysis of such research for Java and JVM languages.

The rapid increase of computational power added a new aspect to create a triangle: object orientation, actor model, and \emph{concurrency}.
Concurrency makes it harder to verify programs in terms of correctness of runtime behavior~\cite{Herlihy:1990:linear,johnsen:history,agha:predictive:safety}. 
The main and foremost challenge is how to combine concurrency with object-oriented programming.
Concurrency brings about two aspects in combination of object orientation: \emph{communication} and \emph{execution}.

There are two fundamental approaches towards communication: synchronous and asynchronous. 
In synchronous model, methods of objects are scheduled to be executed in different threads.
The method invocation is synchronous, however, its execution is scheduled to occur in potentially a different thread from that of the caller object.
This model utilizes future values and takes full advantage of object orientation paradigm.
Asynchronous communication is a core element of actor model.
In actor model, every communication is asynchronous. 
A shared or dedicated queue of messages (inbox) is utilized among actors in the system.
The unit of communication is a \emph{message}.
The notion of a message is not bound by a specific definition or object interface.
When an actor receives a message, it may use pattern matching techniques to determine the content of the message.
When the actor completes the processing of a message, it may decide to reply to the message by sending another message.
There are two essential approaches towards execution: run-to-completion (RTC) and co-operative scheduling.
Actor model uses RTC execution style.
When an actor receives a message, the execution cannot be preempted until the message is processed. 
In co-operative scheduling, an object may decide to yield computation power at any time to allow another object make progress.
In contrast with actor model, concurrent active objects~\cite{creol:broch:owe} combine object orientation and actor model. 
Concurrent active objects use asynchronous communication; however, a message is bound by an object-oriented interface.
In addition, co-operation scheduling is the main execution style in concurrent active objects.

ABS~\cite{johnsen2012abs,hahnlehjlssw11} is a modelling language for concurrent objects and distributed systems.
ABS is based on the semantics of the actor model~\cite{actors:agha}.
ABS language provides a set of core features including algebraic data types (ADT), a functional layer, and a concurrency layer.
The ABS concurrency layer uses co-operative scheduling of active objects~\cite{creol:broch:owe}.
ABS semantics is completely expressed in structural operational semantics~\cite{plotkin:sos}.
This allows ABS models to take advantage of various verification methods and static analysis techniques.
% 
Co-operative scheduling in ABS has been additionally extended for real-time scheduling taking priorities and time constraintsinto account~\cite{bjork2013:rtabs,johnsen2012modeling}.
All the above characteristics make the ABS language an attractive choice if it
can be used as a programming language at industrial and business scale.
ABS has mainly developed as a modelling language.

One core challenge is to combine ABS semantics of its concurrency layer with a
programming language.
Considering the mainstream programming languages
\footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}}, Java~\cite{gosling2000java} is one of the most commonly-used.
Since Java 1.5 and the rise of the concurrency API in Java~\cite{jsr166}, substantial effort is spend by the community on how concurrency and distribution can be improved in Java, along the lines of the actor model, as provided by ABS.
However, not all the development on Java libraries and frameworks are based on a rigorous formal semantics. 
Therefore, this gives rise to issues and challenges in terms of correctness, semantic preservation, and reasoning.

The Java language specification (JLS)~\cite{gosling2000java} clearly concludes that Java language is not comptabile and ready to be extended with a functional language that supports algebraic data types.
This has been the focus of research and development to extend Java to a functional language~\cite{odersky1997pizza,henkel2003discovering,nystrom2003polyglot,bracha1998making}.
Scala language~\cite{odersky2004scala} is a result of such relevant research. It is a dynamic, functional, and object-oriented language on JVM.
However, these extensions and libraries lack formal semantics in the context
of concurrency and actor model.

In this thesis, we primarily focus on ABS concurrency layer and deliver its semantics as a Java API.
We argue that the advantage of a correct semantic translation from ABS concurrency layer to its Java equivalent constructs overcomes the additional support of functional layer with \emph{partial} formal semantics.
Our approach enables programmers to model their Java systems in terms of ABS concurrency semantics and then create a correct mapping and implementation from ABS to Java using an API.
In addition, we aim to facilitate ways to actually program ABS models taking advantage of the standard Java libraries.

The core contributions of this thesis target at the intersection of object 
orientation, actor model, and concurrency.
We choose Java as the main programming language and as one of the mainstream 
object-oriented languages. 
We formalize a subset of Java and its concurrency API~\cite{jsr166} to 
facilitate formal verification and reasoning about it.
We create an abstract mapping from a concurrent-object modelling language, 
ABS~\cite{johnsen2012abs}, to programming semantics of Java concurrency (cf. Chapter~\ref{ch:p01:ch02}). 
We provide the formal semantics of the mapping and runtime properties of 
the concurrency layer including deadlines and scheduling policies (cf. Chapter~\ref{ch:p01:ch01}).
We provide an implementation of ABS concurrency layer as a Java API library 
and framework utilizing the latest langauge additions 
in Java 8~\cite{jsr335:lambda:translation} (cf. Chapter~\ref{ch:p02:ch01}).
We design and implement a runtime monitoring framework, JMSeq, to verify the
sequenced execution of methods through code annotatinon in JVM (cf. Chapter~\ref{ch:p03:ch02}). 
In addition, we design a large-scale monitoring system as a real-world 
application; the monitoring system is built with actors and ABS concurrency 
with formal semantics of ABS composed with schedulability 
analysis~\cite{fersman2007task} (cf. Chapter~\ref{ch:p03:ch01}).
